\documentclass[12pt, a4paper]{article}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[a4paper]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{epsfig,amsfonts}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{subfig}
\usepackage{setspace}
\usepackage{hypcap}
%From: https://tex.stackexchange.com/questions/14364/cross-referencing-between-different-files to connect references from external .tex documents; NOTE: to get xr package to work in Overleaf, followed the directions here https://www.overleaf.com/learn/how-to/Cross_referencing_with_the_xr_package_in_Overleaf ; NOTE: seems like when use {xr} and not {xr-hyper}, things work -- get some 'errors' when use xr-hyper, even though within the document looks like things are working fine?
\usepackage{xr}

\title{Differential complex trait architecture across humans: epistasis identified in non-European populations at multiple genomic scales}
\author[1,2]{Michael C. Turchin}
\author[1,3]{Isabella Ting}
\author[1,4,5,*]{Lorin Crawford}
\author[1,2,*,$\dag$]{Sohini Ramachandran}
\affil[1]{Center for Computational Molecular Biology, Brown University}
\affil[2]{Department of Ecology and Evolutionary Biology, Brown University}
\affil[3]{Department of Computer Science, Brown University}
\affil[4]{Department of Biostatistics, Brown University}
\affil[5]{Center for Statistical Science, Brown University}
\affil[$\ast$]{indicates these authors contributed equally}
\affil[$^\dag$]{To whom correspondence should be addressed: sramachandran@brown.edu}

\begin{document}

\maketitle

\begin{abstract}\label{InterPath-Abstract}
Genome-wide association (GWA) studies have identified thousands of significant genetic associations in humans across a number of complex traits. However, the vast majority of these studies use datasets of predominantly European ancestry \cite{Popejoy2016}. It has generally been thought that complex trait genetic architecture should be transferable across populations of different ancestries, but recent work has shown a number of differences in trait architecture across human ancestries, including heterogeneity in both the identified causal variants and estimated effect sizes 
\cite{Martin2017,Wojcik2019}. Here, we report further evidence that complex trait genetic architecture is fundamentally different among human ancestries by jointly leveraging pathway and epistasis analysis.

Under the assumption that a given complex trait may have differential polygenic architectures across human ancestries, we hypothesize that human populations may also be enriched for differences in epistatic effects. However, since polygenic traits tend to have smaller GWA effect sizes, combining variants via pathway analysis may allow us to better reveal these signals. To accomplish this, we extend the concept of identifying marginal epistasis, moving from testing single variants \cite{Crawford2017} to testing groups of variants for nonlinear association with a trait of interest.

We apply our new method to multiple ancestries present in the UK Biobank \cite{Sudlow2015} and explore multiple pathway-related interaction models. Using morphometric traits we find evidence for genome-wide epistasis in African and other non-European populations. We also find evidence that these trends exists on the SNP and gene levels as well. Results also indicate this may be due to increased heterozygosity in non-European populations. This suggests that non-European populations may be well-suited for identifying non-additive effects in human complex trait architecture; this also suggests further evidence that European populations -- predominantly used for epistasis studies -- may indeed be limited and inaccurate proxies for all human ancestries in complex trait research.
\end{abstract}

\section{Introduction}\label{InterPath-Introduction}

Genome-wide association studies (GWAS) have identified over \textcolor{red}{24,000} significant associations between individual genotypes andcomplex human traits. However, the vast majority of these associations were identified in datasets of primarily European ancestry. In a survey of published GWAS from 2009, it was found that only 4\% of the 1.7 million individuals studied were of non-European ancestry (26). This report gained some brief attention, but an initial increase in the representation of non-European ancestries in GWAS has stagnated since 2014 (27,28); this lack of representation is even more disconcerting given the ongoing explosion of available human genomic data. 

It was largely hoped during the GWAS era that human genetic architecture would be consistent across human ancestries, for instance that causal loci and association effect sizes would remain the same among different human populations (26,29-31). However, recent work has begun to directly interrogate this assumption, and the initial findings are indeed showing that genetic architecture is often not the same between human ancestries; studies have shown that not only are association effect size estimates different between ancestries, but at times even causal loci appear to differ as well (32-35). Additionally, polygenic risk score studies, where published effect size estimates are used to predict phenotypes in other populations, are showing that applying European-based estimates to non-European populations consistently produce erratic and nonsensical results (36-39).

It is clear both that non-European populations are incredibly underrepresented in modern GWAS studies and that European populations cannot act as accurate proxies for the rest of the world. Therefore there is an massive need for work that explores the many aspects of complex trait genetic architecture in non-European populations. To help address these needs, we focus here on investigating the importance of epistasis, or genetic interactions, in complex traits across multiple non-European populations. 

Epistasis, despite being a well-established component of complex trait architecture in multiple model organisms (citations), is still regarded cautiously in human genetics (citations). Recent work has begun to show increased evidence of epistasis playing a role in human trait architecture, but still most of this research has been conducted primarily in datasets of European ancestry. Therefore to address this unexplored area of research, we investigated epistasis at multiple genomic scales across multiple European and non-European populations. We extracted both European and non-European subsets from the UK BioBank (UKB) (citation) and identified evidence for epistasis on the SNP-level, pathway-level, and the genome-wide level. We identify that evidence for epistasis varies across human ancestries, and that African populations often have greater evidence for epistasis. We present a novel framework for combining epistasis and pathway analysis and identify pathways that have both population-specific evidence for epistasis as well as pathways that appear to have significant levels of epistasis worldwide. And lastly, we show beginning evidence that one of the driving factors for the higher levels of epistasis apparent in non-European populations may be greater levels of population-wide genetic diversity.


\section{Results}\label{InterPath-Results}

\subsection{Genome Level Epistasis}\label{InterPath-Results-SNPEpistasis}

To investigate epistasis in multiple human populations, we first extracted a variety of human ancestries from the UKB (see Methods). We collected a total of 8 UKB subsets with an average sample size of XXXX among the non-European populations (Supplementary Table 1), including African, Indian, Chinese, and Pakistani subsets. To maximize our sample sizes per subset, we focused on height and body mass index (BMI) as our complex traits of interest. 

The first approach we took to investigate epistasis was to look at genome-wide estimates of phenotypic PVE. Specifically, we were interested in the following variance component model:

\begin{equation}\label{InterPath-GEMMA-Equation-Model}
 := \sum_{\gamma}w_{\gamma} \gamma
\end{equation}

start with ukbiobank information
GEMMA PVE Results

\subsection{SNP Level Epistasis}\label{InterPath-Results-SNPEpistasis}

PLINK and MAPIT results

\subsection{Pathway Level Epistasis}\label{InterPath-Results-SNPGenomeEpistasis}

Explain motivation for moving into InterPath

\subsubsection{InterPath}

Explain InterPath

\subsubsection{InterPath Results}

Show Results
 - number of hits (barplots)
 - overlap between ancestries / phenotypes
 - which pathways are coming up to the top? modes of epistasis being witnessed/seen?

\subsubsection{Replication in British Subsamples}

\subsection{Relationship Between Epistasis and Genetic Diversity}

\section{Discussion}\label{InterPath-Discussion}


\section{Online Methods}\label{InterPath-Online-Methods}

\subsection{UK BioBank Data}

\subsubsection{Population Subsets}
UK BioBank data was applied for and downloaded from XXX. We first grouped and extracted our multiple population subsets by using self-identified ancestry (`African', `British', `Caribbean', `Chinese', `Indian', `Irish', and `Pakistani'). British random subsets of 4,000 and 10,000 individuals, both the original sets and the additional four rounds of replications, were constructed by randomly choosing initial sets of non-overlapping 4,000 and 10,000 individuals; each round of 4,000 and 10,000 individuals were ensured not to overlap with the previous rounds of the same size of subset. 

We then conducted standard quality control (QC) procedures on each of these population subsets. Note that we focused our analyzes on the genotyped chip data throughout the project. First we conducted SNP-level QC by dropping variants that did not meet the following criteria:  minor allele frequency (MAF) >= .01, genotype missingness <= 5\%, and Hardy-Weinberg equilibrium test p-value >= $1\times10^{-6}$. We then conducted individual-level QC via the following steps. Individuals were removed if they did not have genotype missingness >= 5\%. Individuals were also removed if they were a 3\textsuperscript{rd} degree relative or more to someone else in the dataset; specifically the KING relatedness values provided with the UKB data were used to identify related individuals, and one individual from every pair of 3\textsuperscript{rd} degree or more relatives was removed. Individuals were also dropped if they were tagged by any of the following three flags from the UKB data: `het.missing.outliers', `putative.sex.chromosome.aneuploidy', and `excess.relatives'. Lastly, individuals were removed if they were determined to be PCA outliers; this was conducted by running FlashPCA2\cite{Abraham2017} on each population subset separately and identifying individuals that had PC values greater than 7 standard deviations away from the mean for any of the top 6 PCs. 

After this first round of QC procedures, we then proceeded to impute our current population subsets. Since most of the analyses in this project utilized genetic relatedness matrices (GRMs), and variants need to have no missing data for these GRMs, we used imputation primarily to maximize the number of genotyped SNPs that would not be dropped by this stringent threshold (as opposed to using imputation to increase the number of SNPs we were analyzing). To conduct this imputation, we uploaded our population subsets to the University of Michigan Imputation Server\cite{Das2016} and used the following options: Minimac3 for the imputation software, 1000G phase 3 v5 for the reference panel, and Eagle v2.3 for the phasing software. Completed imputed files were then downloaded from the Imputation Server afterwards and treated to further QC steps: imputed variants were intersected back to the original set of genotyped chip variants, variants with imputation quality scores < .3 were removed, and variants that had genotype missingness rates > 0\% were also removed. These steps represent the last of our QC and imputation procedures, and information on the final forms of our UKB population subsets can be found in Supplementary Table (table).

\subsubsection{Phenotypes}

For all the analyses presented in here, height and BMI were our complex traits of interest. Each phenotype was adjusted for age, gender, and assessment center. Following previous pipelines (cite GIANT), each dataset was first divided into male and female subsets. Age was then regressed out within each sex. The resulting residuals were then inverse normalized according to equation \#\# found in the supplement of (citation). These normalized values were then combined back between sexes, and, lastly, assessment center designations \textcolor{red}{(give UKB identifier?)} as detailed by the UKB were regressed out as well. 

\subsubsection{Global Principal Components}

To take into account possible global patterns of population structure, principal components to be used as covariates during analyses were derived by running PCA on our full set of UKB populations. FlashPCA2 was used to run PCA, and as stated, all of our post-QC, post-imputation population files were analyzed together.    

\subsection{PLINK Analyses}

PLINK pairwise epistasis analyses were conducted using PLINK v1.90b4, the `\texttt{-{}-epistasis}' command, and phenotypes that had top 10 global PCs regressed out; PCs were regressed out directly from the phenotypes due to the `\texttt{-{}-epistasis}' function having no explicit form to incorporate covariates. Default values for the `\texttt{-{}-epi1}' and `\texttt{-{}-epi2}' options (0.0001 and 0.01) were kept. 

\subsection{MAPIT Analyses}

\subsection{GEMMA Analyses}

\subsection{InterPath Model}

\subsection{InterPath Analyses}

\subsection{Genetic Diversity Analyses}

\section{URLs}\label{InterPath-URLs}

\section{Acknowledgments}\label{InterPath-Acknowledgments}

\section{Author Contributions}\label{InterPath-Author-Contributions}

\section{Competing Interests}\label{InterPath-Competing-Interests}

\begingroup
\bibliographystyle{apalike}
\setstretch{1.0}
\bibliography{Main}
\endgroup

\clearpage
\section{Supplementary Material}

\captionsetup[figure]{name=Supplementary Figure}
\captionsetup[table]{name=Supplementary Table}
%\setcounter{figure}{0}


\begingroup
%\bibliographystyle{apalike}
\setstretch{1.0}
%\bibliography{Supplementary}
\endgroup


\iffalse

\fi

\end{document}
